---
title: "Sample-wise vs gene-wise permutations in gene set enrichment analysis"
author: "January Weiner^1,†^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

# Abstract

WE show that for low sample numbers, the gene-wise permutations for
calculating GSEA enrichments produce results whichare divergent from both,
results produced by non-randomization based methods and from results
produced using sample-permutation approaches.

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE}
library(tidyverse)
library(limma)
library(GEOquery)
library(DESeq2)
## install bioshmods from github.com/bihealth/bioshmods
#library(bioshmods)
library(tmod)
library(cowplot)
library(stringr)
#library(ggvenn)
library(ggplot2)
#library(ggpval)
#library(eulerr)
library(Biobase)
library(pander)
library(org.Hs.eg.db)
#library(pwr)
theme_set(theme_minimal() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
source("functions.R")
```

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per group/treatment
                # total number of samples = n_stratum * 2 * 2 * 2
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```




```{r pheno_data_download,eval=FALSE}
## this sometimes fails. The RDS of the object is provided
Sys.setenv(VROOM_CONNECTION_SIZE=8*131072)

## with getGEO, we can only get the phenoData
geo <- getGEO("GSE156063")[[1]]
saveRDS(geo, file="GSE156063.geo")
```

```{r pheno_data_cleanup}
geo <- readRDS("GSE156063.geo")
covar <- as(phenoData(geo), "data.frame")

## remove columns with only one value
boring <- map_lgl(covar, ~ length(unique(.x)) == 1)
covar <- covar[ , !boring ] 

## clean up
covar <- covar %>% 
  dplyr::rename(gender = "gender:ch1") %>%
  mutate(disease = gsub(" .*", "", .data[["disease state:ch1"]])) %>%
  mutate(label = description) %>%
  mutate(group = disease) %>%
  arrange(description) %>%
  dplyr::select(all_of(c("title", "label", "gender", "disease", "group")))
```

```{r featuredata_download}
## the counts must be downloaded from GEO separately.
if(!file.exists("GSE156063_swab_gene_counts.csv.gz")) {
  download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE156nnn/GSE156063/suppl/GSE156063_swab_gene_counts.csv.gz",
                "GSE156063_swab_gene_counts.csv.gz")
}

counts <- read_csv("GSE156063_swab_gene_counts.csv.gz")
.tmp <- counts[[1]]
counts <- as.matrix(counts[,-1])
rownames(counts) <- .tmp
#counts <- counts[ , covar$description ]
counts <- counts[ , covar$label ]
lcpm   <- edgeR::cpm(counts, log=TRUE)
#stopifnot(all(colnames(counts) == covar$description))
stopifnot(all(colnames(counts) == covar$label))

annot <- data.frame(ENSEMBL = rownames(counts))

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("ENTREZID"), keytype="ENSEMBL")
annot$ENTREZID <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("SYMBOL"), keytype="ENSEMBL")
annot$SYMBOL <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]

.tmp <- mapIds(org.Hs.eg.db, annot$ENSEMBL, column=c("GENENAME"), keytype="ENSEMBL")
annot$GENENAME <- .tmp[ match(annot$ENSEMBL, names(.tmp)) ]
```


```{r Preparation_of_the_covariates,cache=FALSE}
sel <- covar$group %in% c("no", "SC2")
counts <- counts[ , sel ]
covar  <- covar[ sel, ]
covar$group <- as.character(covar$group)

covar$group.disease <- paste0(covar$group, '_', covar$disease)
rownames(covar) <- covar$label
saveRDS(covar, file="covar.rds")
saveRDS(counts, file="counts.rds")

```

```{r DESeq2,cache=FALSE}
## DESeq2 calculations
## manual cache, since the operation takes a long time
sel <- !is.na(covar$group)

ds2 <- DESeqDataSetFromMatrix(counts[,sel], colData=covar[sel, ], design=~ disease)

ds2_file <- "ds2_cached.rds"
if(!file.exists(ds2_file)) {
  message("Running DESeq2")
  ds2 <- DESeq(ds2)
  saveRDS(ds2, file=ds2_file)
} else {
  message("Reading ds2 from manual cache")
  ds2 <- readRDS(ds2_file)
}
```


```{r montecarlo}
dir <- "montecarlo"
sample_n <- 100
sample_sizes <- c(3, 4, 5, 10, 15, 20, 25, 30)



```





Implementing the fgsea (not really)

 * preparePathwaysAndStats [filtering?]
 * fgseaSimpleImpl 

!fgseaMultilevel (also based on calcGseaStat)

```
    pvals <- fgseaSimpleImpl(pathwayScores, pathwaysSizes, pathwaysFiltered, 
        leadingEdges, permPerProc, seeds, m, stats, BPPARAM, 
        scoreType)

m -> toKeepLength (???)


```







```{r DE_analysis}
library(tidyverse)
library(tmod)
library(fgsea)
library(BiocParallel)

res <- results(ds2, name="disease_SC2_vs_no") %>% as.data.frame() %>%
  rownames_to_column("ENSEMBL") %>% left_join(annot, by="ENSEMBL") %>%
  arrange(pvalue)

tmod2List <- function(mset) {

  ret <- map(mset$gs2gv, ~ {
               mset$gv[ .x ]
                })
  names(ret) <- mset$gs$ID
  return(ret)
}

data(tmod)
tmod_fgsea <- tmod2List(tmod)
stats <- set_names(res$log2FoldChange, res$SYMBOL)
stats <- stats[ !duplicated(names(stats)) ]
fgsea_res <- fgseaMultilevel(tmod_fgsea, stats)

pp <- fgsea:::preparePathwaysAndStats(tmod_fgsea, stats, minSize=1,
                                      maxSize=Inf, gseaParam=1,
                                      scoreType="std")

fgsea_res_2 <- map(pp$filtered, ~ {
  fgsea:::calcGseaStat(.x, stats=pp$stats, returnLeadingEdge=T,
                     scoreType="std")
                                      })

gseaStatRes <- do.call(rbind,
                       lapply(pp$filtered,
                              calcGseaStat,
                              stats             = pp$stats,
                              returnLeadingEdge = TRUE,
                              scoreType         = "std"))

leadingEdges <- mapply("[", list(names(pp$stats)), gseaStatRes[, "leadingEdge"], SIMPLIFY = FALSE)
pathwayScores <- unlist(gseaStatRes[, "res"])

foo <- do.call(rbind, fgsea_res_2)

lead_edges <- mapply("[", list(names(pp$stats)), foo[, "leadingEdge"], SIMPLIFY = FALSE)

m <- length(pp$filtered)

simpleFgseaRes <- fgsea:::fgseaSimpleImpl(pathwayScores=pathwayScores,
	pathwaysSizes=pp$sizes,
  pathwaysFiltered=pp$filtered, 
	leadingEdges=leadingEdges,
  permPerProc=1000, 
	seeds=1234, 
  toKeepLength=m,
  stats=pp$stats, BPPARAM=SerialParam(), scoreType="std")


```

```{r}
pp <- fgsea:::preparePathwaysAndStats(tmod_fgsea, stats, minSize=1,
                                      maxSize=Inf, gseaParam=1, 
                                      scoreType="std")

```



```{r}
sessioninfo::session_info(to_file=TRUE)
```




## Bibliography

